---
title: "Chapter 12: Creating Machine Learning Datasets Using SQL"
output: html_notebook
---

# Setup Database

```{r setup}
library(pacman)
p_load(dplyr, DBI)

# Create a database table in memory
farmers.db <- 
  dbConnect(RMariaDB::MariaDB(), 
            group = "/Users/nickjenkins/Documents/Workshops & Conferences/SQL for Data Scientists/SQL for Data Scientists support files/FarmersMarketDatabase.sql",
            username = "root",
            password = "Hockeydude25",
            host = "localhost",
            port = 3306)
```

# Creating Machine Learning Datasets Using SQL

In this chapter, we'll discuss the development of datasets for two types of algorithms: classification and time series models.

A *binary classification model* predicts whether a record belongs to one category or another. In order to make predictions, the model needs to be trained. Binary classifiers are a type of *supervised learning* model, which means thy are trained by passing example rows of data labeled with each of the possible outcomes into the algorithm, so it can detect patterns and identify characteristics that are more strongly associated with one result or the other.

A *time series model* preforms statistical operations on a series of measurements over time to forecast what the measurement might be at some point in time in the future.

# Datasets for Time Series Models

The simplest time series forecast uses a single variable measured over specified time intervals to predict the value of that same variable at a future point in time. For example, a model to predict the high temperature in a location tomorrow could have a dataset with year's worth of daily high temperatures measured at that location. The dataset would have one row per day, with one column for the date, and another for the daily high temperature measured. A time series algorithm could detect seasonal temperature patterns, long-term trends, and the most recent daily high temperatures to predict what the high temperature might be tomorrow.

Let's create a dataset that allows us to plot a time series of farmer's market sales per week.

```{sql connection = farmers.db, output.var = "ts.data"}
SELECT
  MIN(cp.market_date) AS first_market_date_of_week,
  ROUND(SUM(cp.quantity * cp.cost_to_customer_per_qty), 2) AS weekly_sale
FROM farmers_market.customer_purchases AS cp
LEFT JOIN farmers_market.market_date_info AS md
  ON cp.market_date = md.market_date
GROUP BY md.market_year, md.market_week
ORDER BY md.market_year, md.market_week
```

R also has packages for exponential smoothing models, so we'll use R to generate a sales forecast for the eight weeks beyond the last date in our dataset.

```{r}
p_load(tidyverse, modeltime, lubridate, tidymodels, xgboost, timetk)

# series plot
ts.data %>% 
  filter(date >= "2020-03-01" & date <= "2020-11-22") %>% 
  ggplot(aes(x = first_market_date_of_week, y = weekly_sale)) +
  geom_line() +
  scale_y_continuous(limits = c(580, 1250), 
                     breaks = seq(from = 600, to = 1250, by = 50)) +
  scale_x_date(date_breaks = "1 weeks") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 90))

f.data <- 
  ts.data %>% 
  filter(date >= "2020-03-01" & date <= "2020-11-22")

# split the data
splits <- initial_time_split(f.data, prop = 0.8)

# exponential smoothing model
exp.fit <- 
  exp_smoothing() %>% 
  set_engine(engine = "ets") %>% 
  fit(weekly_sale ~ first_market_date_of_week, data = training(splits))

# ARIMA model
arima.fit <- 
  arima_reg() %>% 
  set_engine(engine = "auto_arima") %>% 
  fit(weekly_sale ~ first_market_date_of_week, data = training(splits))

# create a model table
model.tbl <- modeltime_table(exp.fit, arima.fit)

# calibrate the model to the testing set
calibration.table <-
  model.tbl %>% 
  modeltime_calibrate(new_data = testing(splits))

# visualize the forecast test
calibration.table %>% 
  modeltime_forecast(new_data = testing(splits),
                     actual_data = f.data) %>% 
  plot_modeltime_forecast(.interactive = FALSE)

# forecast forward
refit <- 
  calibration.table %>% 
  modeltime_refit(data = f.data)

refit %>% 
  modeltime_forecast(h = "3 weeks", actual_data = f.data) %>% 
  plot_modeltime_forecast(.interactive = FALSE)
```

# Datasets for Binary Classification

Classification algorithms detect patterns in training datasets, which contain example records from the past that fall into known categories, and use those detected patterns to then categorize new data. Binary classification algorithms categorize inputs into one of two outcomes, which are determined according to the purpose of the model and the available data.

## Creating the Dataset

As an example, we'll build a dataset that could be used to train a model that can predict an answer to the question "Will this customer who just made a purchase return to make another purchase within the next month?" The binary target variable is "makes another purchase within 30 days." In the following queries, the target variable `purchased_again_within_30_days` represents whether the customer returned within 30 days of making a purchase, without considering the calendar month.

```{sql connection = farmers.db}
WITH
  customer_markets_attended AS
  (
    SELECT DISTINCT
      customer_id,
      market_date
    FROM farmers_market.customer_purchases
    ORDER BY customer_id, market_date
  )

SELECT 
  cp.market_date,
  cp.customer_id,
  SUM(cp.quantity * cp.cost_to_customer_per_qty) AS purchase_total,
  COUNT(DISTINCT cp.vendor_id) AS vendors_patronized,
  COUNT(DISTINCT cp.product_id) AS different_products_purchased,
  (SELECT MIN(cma.market_date)
  FROM customer_markets_attended AS cma
  WHERE cma.customer_id = cp.customer_id
  AND cma.market_date > cp.market_date
  GROUP BY cma.customer_id) AS customer_next_market_date,
  DATEDIFF(
  (SELECT MIN(cma2.market_date)
  FROM customer_markets_attended AS cma2
  WHERE cma2.customer_id = cp.customer_id
  AND cma2.market_date > cp.market_date
  GROUP BY cma2.customer_id),
  cp.market_date) AS days_until_customer_next_market_date,
  CASE WHEN
  DATEDIFF(
  (SELECT MIN(cma3.market_date)
  FROM customer_markets_attended AS cma3
  WHERE cma3.customer_id = cp.customer_id
  AND cma3.market_date > cp.market_date
  GROUP BY cma3.customer_id),
  cp.market_date) <= 30
  THEN 1
  ELSE 0 END AS purchased_again_within_30_days
FROM farmers_market.customer_purchases AS cp
GROUP BY cp.customer_id, cp.market_date
ORDER BY cp.customer_id, cp.market_date
```

-   The `purchase_total` column multiplies the quantity and cost of each item purchased and summing that up to get the total spent by each customer at each market date.

-   The `vendors_patronized` column is a distinct count of how many different vendors the customer made purchases from that day

-   The `different_products_purchased` column is a distinct count of how many different kinds of products the customer purchased

-   The `customer_next_market_date` is generated by a subquery that references our CTE. Look at the code inside the parentheses after `different_products_purchased` and before `customer_next_market_date`. This subquery selects the minimum market date a customer attended, which occurs after the current row's `market_date` value. The `WHERE` clause of this subquery matches the subquery's `customer_id` with the main query's `customer_id`.

-   The `days_until_customer_next_market_date` uses the same subquery but calculates the difference between the current row's date and that calculated next market date.

-   The last column `purchased_again_within_30_days` uses the same calculation and wraps it inside a `CASE` statement that returns a binary flag indicating whether that next purchase was made within 30 days.

## Expanding the Feature Set

Let's add some columns that indicate which vendors each customer shopped at on each market day and flip the `days_until_customer_next_market_date` calculation to indicate how long it's been since the customer last shopped before the visit represented by the row:

```{sql connection = farmers.db}
WITH
  customer_markets_attended AS
  (
    SELECT DISTINCT
      customer_id,
      market_date
    FROM farmers_market.customer_purchases
    ORDER BY customer_id, market_date
  )

SELECT 
  cp.market_date,
  cp.customer_id,
  SUM(cp.quantity * cp.cost_to_customer_per_qty) AS purchase_total,
  COUNT(DISTINCT cp.vendor_id) AS vendors_patronized,
  MAX(CASE WHEN cp.vendor_id = 7 THEN 1 ELSE 0 END) AS purchased_from_vendor_7,
  MAX(CASE WHEN cp.vendor_id = 8 THEN 1 ELSE 0 END) AS purchased_from_vendor_8,
  COUNT(DISTINCT cp.product_id) AS different_products_purchased,
  DATEDIFF(cp.market_date,
  (SELECT MAX(cma.market_date)
  FROM customer_markets_attended AS cma
  WHERE cma.customer_id = cp.customer_id
  AND cma.market_date > cp.market_date
  GROUP BY cma.customer_id)) AS days_since_last_customer_market_date,
  CASE WHEN
  DATEDIFF(
  (SELECT MIN(cma3.market_date)
  FROM customer_markets_attended AS cma3
  WHERE cma3.customer_id = cp.customer_id
  AND cma3.market_date > cp.market_date
  GROUP BY cma3.customer_id),
  cp.market_date) <= 30 THEN 1 ELSE 0 END AS purchased_again_within_30_days
FROM farmers_market.customer_purchases AS cp
GROUP BY cp.customer_id, cp.market_date
ORDER BY cp.customer_id, cp.market_date
```

Now let's add a column that represents the number of times each customer has shopped at a given farmer's market. To do this, we will move the `ROW_NUMBER()` into the `WHERE` clause to return the same count for `customer_markets_attended_count` as a version that has it in the main query even if the main query is filtered to a date range that doesn't include a customer's entire purchase history:

```{sql connection = farmers.db}
WITH
  customer_markets_attended AS
  (
    SELECT DISTINCT
      customer_id,
      market_date,
      ROW_NUMBER() OVER (PARTITION BY customer_id ORDER BY market_date) AS market_count
    FROM farmers_market.customer_purchases
    ORDER BY customer_id, market_date
  )

SELECT 
  cp.market_date,
  cp.customer_id,
  (SELECT MAX(market_count)
  FROM customer_markets_attended AS CMA
  WHERE cma.customer_id = cp.customer_id
  AND cma.market_date <= cp.market_date) AS customer_markets_attended_count
FROM farmers_market.customer_purchases AS cp
GROUP BY cp.customer_id, cp.market_date
ORDER BY cp.customer_id, cp.market_date
```

## Feature Engineering

The process of creating different input values that might be helpful to the prediction algorithm is called *feature engineering*. One important factor when engineering features is that each of these feature values is only what would be knowable as of the date represented by the row. We want to train the model on examples of customers with a variety of traits as of specific points in time that can be correlated with a specific outcome or target variable relative to that time.

# Exercises

1.  Add a column to the final query in the chapter that counts how many markets were attended by each customer in the past 14 days.

    ```{sql connection = farmers.db}
    WITH
      customer_markets_attended AS
      (
        SELECT DISTINCT
          customer_id,
          market_date,
          ROW_NUMBER() OVER (PARTITION BY customer_id ORDER BY market_date) AS market_count
        FROM farmers_market.customer_purchases
        ORDER BY customer_id, market_date
      )

    SELECT 
      cp.market_date,
      cp.customer_id,
      EXTRACT(MONTH FROM cp.market_date) AS market_month,
      SUM(cp.quantity * cp.cost_to_customer_per_qty) AS purchase_total,
      COUNT(DISTINCT cp.vendor_id) AS vendors_patronized,
      MAX(CASE WHEN cp.vendor_id = 7 THEN 1 ELSE 0 END) AS purchased_from_vendor_7,
      MAX(CASE WHEN cp.vendor_id = 8 THEN 1 ELSE 0 END) AS purchased_from_vendor_8,
      COUNT(DISTINCT cp.product_id) AS different_products_purchased,
      DATEDIFF(cp.market_date,
      (SELECT MAX(cma.market_date)
      FROM customer_markets_attended AS cma
      WHERE cma.customer_id = cp.customer_id
      AND cma.market_date > cp.market_date
      GROUP BY cma.customer_id)) AS days_since_last_customer_market_date,
      (SELECT MAX(market_count)
      FROM customer_markets_attended AS cma
      WHERE cma.customer_id = cp.customer_id
      AND cma.market_date <= cp.market_date) AS customer_markets_attended_count,
      (SELECT COUNT(market_date)
      FROM customer_markets_attended AS cma
      WHERE cma.customer_id = cp.customer_id
      AND cma.market_date < cp.market_date
      AND DATEDIFF(cp.market_date, cma.market_date) <= 30) AS
        customer_markets_attended_30days_count,
      (SELECT COUNT(market_date)
      FROM customer_markets_attended AS cma
      WHERE cma.customer_id = cp.customer_id
      AND cma.market_date < cp.market_date
      AND DATEDIFF(cp.market_date, cma.market_date) <= 14) AS
        customer_markets_attended_14days_count,
      CASE WHEN
      DATEDIFF(
      (SELECT MIN(cma3.market_date)
      FROM customer_markets_attended AS cma3
      WHERE cma3.customer_id = cp.customer_id
      AND cma3.market_date > cp.market_date
      GROUP BY cma3.customer_id),
      cp.market_date) <= 30 THEN 1 ELSE 0 END AS purchased_again_within_30_days
    FROM farmers_market.customer_purchases AS cp
    GROUP BY cp.customer_id, cp.market_date
    ORDER BY cp.customer_id, cp.market_date
    ```

2.  Add a column to the final query in the chapter that contains a 1 if the customer purchased an item that cost over \$10, and a 0 if not. HINT: The calculation will follow the same form as the `purchased_from_vendor_x` flags.

    ```{sql connection = farmers.db}
    WITH
      customer_markets_attended AS
      (
        SELECT DISTINCT
          customer_id,
          market_date,
          ROW_NUMBER() OVER (PARTITION BY customer_id ORDER BY market_date) AS market_count
        FROM farmers_market.customer_purchases
        ORDER BY customer_id, market_date
      )

    SELECT 
      cp.market_date,
      cp.customer_id,
      EXTRACT(MONTH FROM cp.market_date) AS market_month,
      SUM(cp.quantity * cp.cost_to_customer_per_qty) AS purchase_total,
      COUNT(DISTINCT cp.vendor_id) AS vendors_patronized,
      MAX(CASE WHEN cp.vendor_id = 7 THEN 1 ELSE 0 END) AS purchased_from_vendor_7,
      MAX(CASE WHEN cp.vendor_id = 8 THEN 1 ELSE 0 END) AS purchased_from_vendor_8,
      COUNT(DISTINCT cp.product_id) AS different_products_purchased,
      MAX(CASE WHEN cp.cost_to_customer_per_qty > 10 THEN 1 ELSE 0 END) AS
        purchased_over_ten,
      DATEDIFF(cp.market_date,
      
      (SELECT MAX(cma.market_date)
      FROM customer_markets_attended AS cma
      WHERE cma.customer_id = cp.customer_id
      AND cma.market_date > cp.market_date
      GROUP BY cma.customer_id)) AS days_since_last_customer_market_date,
      
      (SELECT MAX(market_count)
      FROM customer_markets_attended AS cma
      WHERE cma.customer_id = cp.customer_id
      AND cma.market_date <= cp.market_date) AS customer_markets_attended_count,
      
      (SELECT COUNT(market_date)
      FROM customer_markets_attended AS cma
      WHERE cma.customer_id = cp.customer_id
      AND cma.market_date < cp.market_date
      AND DATEDIFF(cp.market_date, cma.market_date) <= 30) AS
        customer_markets_attended_30days_count,
      
      (SELECT COUNT(market_date)
      FROM customer_markets_attended AS cma
      WHERE cma.customer_id = cp.customer_id
      AND cma.market_date < cp.market_date
      AND DATEDIFF(cp.market_date, cma.market_date) <= 14) AS
        customer_markets_attended_14days_count,
      
      CASE WHEN
      DATEDIFF(
      (SELECT MIN(cma3.market_date)
      FROM customer_markets_attended AS cma3
      WHERE cma3.customer_id = cp.customer_id
      AND cma3.market_date > cp.market_date
      GROUP BY cma3.customer_id),
      cp.market_date) <= 30 THEN 1 ELSE 0 END AS purchased_again_within_30_days

    FROM farmers_market.customer_purchases AS cp
    GROUP BY cp.customer_id, cp.market_date
    ORDER BY cp.customer_id, cp.market_date
    ```

3.  Let's say that the farmer's market started a customer reward program that gave customers a market goods gift basket and branded reusable market bag when they had spent at least \$200 total. Create a flag field that indicates whether the customer has reached this loyal customer status. HINT: One way to accomplish this involves modifying the CTE (`WITH` clause) to include purchase totals, and adding a column to the main query with a similar structure to the one that calculates `customer_markets_attended_count` to calculate a running total spent.

    ```{sql connection = farmers.db}
    WITH
      customer_markets_attended AS
      (
        SELECT DISTINCT
          customer_id,
          market_date,
          SUM(quantity * cost_to_customer_per_qty) AS purchase_total,
          ROW_NUMBER() OVER (PARTITION BY customer_id ORDER BY market_date) AS
            market_count
        FROM farmers_market.customer_purchases
        GROUP BY customer_id, market_date
        ORDER BY customer_id, market_date
      )

    SELECT 
      cp.market_date,
      cp.customer_id,
      EXTRACT(MONTH FROM cp.market_date) AS market_month,
      SUM(cp.quantity * cp.cost_to_customer_per_qty) AS purchase_total,
      COUNT(DISTINCT cp.vendor_id) AS vendors_patronized,
      MAX(CASE WHEN cp.vendor_id = 7 THEN 1 ELSE 0 END) AS purchased_from_vendor_7,
      MAX(CASE WHEN cp.vendor_id = 8 THEN 1 ELSE 0 END) AS purchased_from_vendor_8,
      COUNT(DISTINCT cp.product_id) AS different_products_purchased,
      MAX(CASE WHEN cp.cost_to_customer_per_qty > 10 THEN 1 ELSE 0 END) AS
        purchased_over_ten,
      DATEDIFF(cp.market_date,
      
      (SELECT MAX(cma.market_date)
      FROM customer_markets_attended AS cma
      WHERE cma.customer_id = cp.customer_id
      AND cma.market_date > cp.market_date
      GROUP BY cma.customer_id)) AS days_since_last_customer_market_date,
      
      (SELECT MAX(market_count)
      FROM customer_markets_attended AS cma
      WHERE cma.customer_id = cp.customer_id
      AND cma.market_date <= cp.market_date) AS customer_markets_attended_count,
      
      (SELECT COUNT(market_date)
      FROM customer_markets_attended AS cma
      WHERE cma.customer_id = cp.customer_id
      AND cma.market_date < cp.market_date
      AND DATEDIFF(cp.market_date, cma.market_date) <= 30) AS
        customer_markets_attended_30days_count,
      
      (SELECT COUNT(market_date)
      FROM customer_markets_attended AS cma
      WHERE cma.customer_id = cp.customer_id
      AND cma.market_date < cp.market_date
      AND DATEDIFF(cp.market_date, cma.market_date) <= 14) AS
        customer_markets_attended_14days_count,
      
      CASE WHEN
      DATEDIFF(
      (SELECT MIN(cma3.market_date)
      FROM customer_markets_attended AS cma3
      WHERE cma3.customer_id = cp.customer_id
      AND cma3.market_date > cp.market_date
      GROUP BY cma3.customer_id),
      cp.market_date) <= 30 THEN 1 ELSE 0 END AS purchased_again_within_30_days,
      
      CASE WHEN
      (SELECT SUM(purchase_total)
      FROM customer_markets_attended AS cma
      WHERE cma.customer_id = cp.customer_id
      AND cma.market_date <= cp.market_date) > 200
      THEN 1 ELSE 0 END AS spent_over_200

    FROM farmers_market.customer_purchases AS cp
    GROUP BY cp.customer_id, cp.market_date
    ORDER BY cp.customer_id, cp.market_date
    ```
