---
title: "Chapter 11: More Advanced Query Structures"
output: html_notebook
---

# Setup Database

```{r setup}
library(pacman)
p_load(dplyr, DBI)

# Create a database table in memory
farmers.db <- 
  dbConnect(RMariaDB::MariaDB(), 
            group = "/Users/nickjenkins/Documents/Workshops & Conferences/SQL for Data Scientists/SQL for Data Scientists support files/FarmersMarketDatabase.sql",
            username = "root",
            password = "Hockeydude25",
            host = "localhost",
            port = 3306)
```

# More Advanced Query Structures

# `UNION`s

Using a `UNION` you can combine any two queries that result in the same number of columns with the same data types. They syntax is simple: write two queries with the same number and type of fields, and put a `UNION` keyword between them:

```{sql connection = farmers.db}
SELECT 
  market_year, 
  MIN(market_date) AS first_market_date
FROM farmers_market.market_date_info
WHERE market_year = "2019"

UNION

SELECT 
  market_year, 
  MIN(market_date) AS first_market_date
FROM farmers_market.market_date_info
WHERE market_year = "2020"
```

For a more complex example combining CTEs and `UNION`s, we'll build a report that shows the products with the largest quantities available to each market: the bulk product with the largest weight available, and the unit product with the highest count available:

```{sql connection = farmers.db}
WITH product_quantity_by_date AS
(
  SELECT
    vi.market_date,
    vi.product_id,
    p.product_name,
    p.product_qty_type,
    SUM(vi.quantity) AS total_quantity_available
  FROM farmers_market.vendor_inventory AS vi
  LEFT JOIN farmers_market.product AS p
    ON vi.product_id = p.product_id
  GROUP BY market_date, product_id, product_name, product_qty_type
)

SELECT *
FROM
(
  SELECT
    market_date,
    product_id,
    product_name,
    total_quantity_available,
    product_qty_type,
    RANK() OVER (PARTITION BY market_date ORDER BY total_quantity_available DESC) AS quantity_rank
  FROM product_quantity_by_date
  WHERE product_qty_type = "unit"
  
  UNION
  
  SELECT
    market_date, 
    product_id,
    product_name,
    total_quantity_available,
    product_qty_type,
    RANK() OVER (PARTITION BY market_date ORDER BY total_quantity_available DESC) AS quantity_rank
  FROM product_quantity_by_date
  WHERE product_qty_type = "lbs"
) x

WHERE x.quantity_rank = 1
ORDER BY market_date
```

The `WITH` statement at the top of this query totals up the quantity of each product that is available at each market from the `vendor_invenroty` table, and joins in helpful information from the product table such as product name and the type of quantity, which has `product_aty_type` values of "lbs" or "unit."

The inner part of the bottom query contains two different queries of the same view created in the CTE, `product_quantity_by_date`, `UNION`ed together. Each ranks the information available in the CTE by `total_quantity_available` as well as returning all of the available fields.

Here is the same result without the `UNION` statement:

```{sql connection = farmers.db}
WITH product_quantity_by_date AS
(
  SELECT
    vi.market_date,
    vi.product_id,
    p.product_name,
    p.product_qty_type,
    SUM(vi.quantity) AS total_quantity_available
  FROM farmers_market.vendor_inventory AS vi
  LEFT JOIN farmers_market.product AS p
    ON vi.product_id = p.product_id
  GROUP BY market_date, product_id, product_name, product_qty_type
),
rank_by_qty_type AS
(
  SELECT
    market_date, 
    product_id,
    product_name,
    total_quantity_available,
    product_qty_type,
    RANK() OVER (PARTITION BY market_date, product_qty_type ORDER BY 
    total_quantity_available DESC) AS quantity_rank
  FROM product_quantity_by_date
)

SELECT * 
FROM rank_by_qty_type
WHERE quantity_rank = 1
ORDER BY market_date
```

We were able to accomplish the same result without the `UNION` by partitioning by both the `market_date` and `product_qty_type` in the `RANK()` function, resulting in a ranking for each date and quantity type.

# Self-Join to Determine To-Date Maximum

A *self-join* in SQL is when a table is joined to itself in order to compare rows to one another.

Let's say we wanted to show an aggregate metric changing over time, comparing each value to all previous values. One reasons you might want to compare a value to all previous values it to create a "record high to-date" indicator.

As an example, we create a report showing whether the total sales on each market date were the highest for any market to-date. If we were always looking at data filtered to dates before a selected date, we could simply use the `SUM()` and `MAX()` functions to determine the highest total sales for the given date range. But, we want to determine whether there is any previous date that has a higher sales total than the "current" row we're looking at, and we can use a self-join to do that comparison. First, we'll need to summarize the sales by `market_date`, which we have done previously. We'll put this query into a CTE and alias it `sales_per_market_date`.

```{sql fig-11.2, connection = farmers.db}
WITH sales_per_market_date AS
(
  SELECT
    market_date,
    ROUND(SUM(quantity * cost_to_customer_per_qty), 2) AS sales
  FROM farmers_market.customer_purchases
  GROUP BY market_date
  ORDER BY market_date
)

SELECT *
FROM sales_per_market_date
LIMIT 10
```

We can select data from the "table" twice. We do this by joining the table to itself on every row where the row has a date that occurred prior to the "current" row's date.

```{sql fig-11.3, connection = farmers.db}
WITH sales_per_market_date AS
(
  SELECT
    market_date,
    ROUND(SUM(quantity * cost_to_customer_per_qty), 2) AS sales
  FROM farmers_market.customer_purchases
  GROUP BY market_date
  ORDER BY market_date
)

SELECT *
FROM sales_per_market_date AS cm
LEFT JOIN sales_per_market_date AS pm
  ON pm.market_date < cm.market_date
WHERE cm.market_date = "2019-04-13"
```

Now we'll use a `MAX()` function on the `pm.sales` field and `GROUP BY` `cm.market_date` to get previous highest sales value.

```{sql fig-11.4, connection = farmers.db}
WITH sales_per_market_date AS
(
  SELECT
    market_date,
    ROUND(SUM(quantity * cost_to_customer_per_qty), 2) AS sales
  FROM farmers_market.customer_purchases
  GROUP BY market_date
  ORDER BY market_date
)

SELECT 
  cm.market_date,
  cm.sales,
  MAX(pm.sales) AS previous_max_sales
FROM sales_per_market_date AS cm
LEFT JOIN sales_per_market_date AS pm
  ON pm.market_date < cm.market_date
WHERE cm.market_date = "2019-04-13"
GROUP BY cm.market_date, cm.sales
```

We can now remove the date filter in the `WHERE` clause to get the `previous_max` sales for each date. We can also use a `CASE` statement to create a flag field that indicates whether the current sales are higher than the previous maximum sales, indicating whether each row's `market_date` set a sales record as of that date.

```{sql fig-11.5, connection = farmers.db}
WITH sales_per_market_date AS
(
  SELECT
    market_date,
    ROUND(SUM(quantity * cost_to_customer_per_qty), 2) AS sales
  FROM farmers_market.customer_purchases
  GROUP BY market_date
  ORDER BY market_date
)

SELECT 
  cm.market_date,
  cm.sales,
  MAX(pm.sales) AS previous_max_sales,
  CASE WHEN cm.sales > MAX(pm.sales) THEN "YES" ELSE "NO"
  END sales_record_set
FROM sales_per_market_date AS cm
LEFT JOIN sales_per_market_date AS pm
  ON pm.market_date < cm.market_date
GROUP BY cm.market_date, cm.sales
```

# Counting New vs. Returning Customers by Week

Another common report has to do with summarizing customers by time period. The manager of the farmer's market might want to monitor how many customers are visiting the market per week, and how many of those are new, making a purchase for the first time.

One way to determine if a customer is new is to compare the purchase date to the minimum purchase date per customer.

```{sql fig-11.6, connection = farmers.db}
SELECT
  DISTINCT customer_id,
  market_date,
  MIN(market_date) OVER(PARTITION BY cp.customer_id) AS first_purchase_date
FROM farmers_market.customer_purchases AS CP
```

Now we can put this query inside a `WITH` clause and query its results with some calculations added. We also need to join it to the `market_date_info` table to get the year and week of each `market_date`.

```{sql fig-11.7, connection = farmers.db}
WITH customer_markets_attended AS
(
  SELECT DISTINCT
    customer_id,
    market_date,
    MIN(market_date) OVER(PARTITION BY cp.customer_id) AS first_purchase_date
  FROM farmers_market.customer_purchases AS CP
)

SELECT
  md.market_year,
  md.market_week,
  COUNT(customer_id) AS customer_visit_count,
  COUNT(DISTINCT customer_id) AS distinct_customer_count
FROM customer_markets_attended AS cma
LEFT JOIN farmers_market.market_date_info AS md
  ON cma.market_date = md.market_date
GROUP BY md.market_year, md.market_week
ORDER BY md.market_year, md.market_week
```

We could have produced this result with a simpler query, but now we have the data in a form that facilitates further analysis. We want to get a count of new customers per week, so let's add a column displaying what percent of each week's customers are new.

```{sql fig-11.8, connection = farmers.db}
WITH customer_markets_attended AS
(
  SELECT DISTINCT
    customer_id,
    market_date,
    MIN(market_date) OVER(PARTITION BY cp.customer_id) AS first_purchase_date
  FROM farmers_market.customer_purchases AS CP
)

SELECT
  md.market_year,
  md.market_week,
  COUNT(customer_id) AS customer_visit_count,
  COUNT(DISTINCT customer_id) AS distinct_customer_count,
  COUNT(
  DISTINCT
  CASE WHEN cma.market_date = cma.first_purchase_date
  THEN customer_id
  ELSE NULL
  END
  ) AS new_customer_count,
  COUNT(
  DISTINCT
  CASE WHEN cma.market_date = cma.first_purchase_date
  THEN customer_id
  ELSE NULL
  END) / COUNT(DISTINCT customer_id)
  AS new_customer_pct
FROM customer_markets_attended AS cma
LEFT JOIN farmers_market.market_date_info AS md
  ON cma.market_date = md.market_date
GROUP BY md.market_year, md.market_week
ORDER BY md.market_year, md.market_week
```

# Exercises

1.  Starting with the query associated with Figure 11.5, put the larger SELECT statement in a second CTE, and write a query that queries from its results to display the current record sales and associated market date. Can you think of another way to generate the same results?

    ```{sql ex1, connection = farmers.db}
    WITH sales_per_market_date AS
    (
      SELECT
        market_date,
        ROUND(SUM(quantity * cost_to_customer_per_qty), 2) AS sales
      FROM farmers_market.customer_purchases
      GROUP BY market_date
      ORDER BY market_date
    ), sales_records AS
    (
      SELECT 
        cm.market_date,
        cm.sales,
        MAX(pm.sales) AS previous_max_sales,
        CASE WHEN cm.sales > MAX(pm.sales) THEN "YES" ELSE "NO"
        END sales_record_set
      FROM sales_per_market_date AS cm
      LEFT JOIN sales_per_market_date AS pm
        ON pm.market_date < cm.market_date
      GROUP BY cm.market_date, cm.sales
    )

    SELECT
      market_date,
      sales
    FROM sales_records
    WHERE sales_record_set = "YES"
    ORDER BY market_date DESC
    LIMIT 1
    ```

2.  Modify the "New vs. Returning Customers Per Week" report (associated with Figure 11.8) to summarize the counts by vendor by week.

    ```{sql ex2, connection = farmers.db}
    WITH customer_markets_attended AS
    (
      SELECT DISTINCT
        customer_id,
        vendor_id,
        market_date,
        MIN(market_date) OVER(PARTITION BY cp.customer_id, cp.vendor_id) AS first_purchase_date
      FROM farmers_market.customer_purchases AS cp
    )

    SELECT
      md.market_year,
      md.market_week,
      cmv.vendor_id,
      COUNT(customer_id) AS customer_visit_count,
      COUNT(DISTINCT customer_id) AS distinct_customer_count,
      COUNT(
      DISTINCT
      CASE WHEN cmv.market_date = cmv.first_purchase_date
      THEN customer_id
      ELSE NULL
      END
      ) AS new_customer_count,
      COUNT(
      DISTINCT
      CASE WHEN cmv.market_date = cmv.first_purchase_date
      THEN customer_id
      ELSE NULL
      END) / COUNT(DISTINCT customer_id)
      AS new_customer_pct
    FROM customer_markets_attended AS cmv
    LEFT JOIN farmers_market.market_date_info AS md
      ON cmv.market_date = md.market_date
    GROUP BY md.market_year, md.market_week, cmv.vendor_id
    ORDER BY md.market_year, md.market_week, cmv.vendor_id
    ```

3.  Using a UNION , write a query that displays the market dates with the highest and lowest total sales.

    ```{sql ex3, connection = farmers.db}
    WITH sales_per_market_date AS
    (
      SELECT
        market_date,
        ROUND(SUM(quantity * cost_to_customer_per_qty), 2) AS sales
      FROM farmers_market.customer_purchases
      GROUP BY market_date
    ), market_dates_ranked_by_sales AS
    (
      SELECT
        market_date,
        sales,
        RANK() OVER (ORDER BY sales) AS sales_rank_asc,
        RANK() OVER (ORDER BY sales) AS sales_rank_desc
      FROM sales_per_market_date
    )

    SELECT 
      market_date,
      sales,
      sales_rank_asc AS sales_rank
    FROM market_dates_ranked_by_sales
    WHERE sales_rank_asc = 1

    UNION

    SELECT 
      market_date,
      sales,
      sales_rank_desc AS sales_rank
    FROM market_dates_ranked_by_sales
    WHERE sales_rank_desc = 1
    ```

    The solution in R:

    First, pull the data:

    ```{sql connection = farmers.db, output.var = "purchases"}
    SELECT 
      market_date,
      ROUND(SUM(quantity * cost_to_customer_per_qty), 2) AS sales
    FROM farmers_market.customer_purchases
    GROUP BY market_date
    ```

    Now R:

    ```{r}
    library(tidyverse)
    purchases %>% 
      filter(sales == max(sales) | sales == min(sales))
    ```
